\chapter{Homotopy}

\section{Homotopy of Maps}

\bx
Define homotopy for maps in the category $\mathcal{T}_{(2)}$ of pairs.
\ex

\bs
Intuitively, a homotopy between maps $f,g\in\map_{(2)}((X,A),(Y,B))$ should be a path $H\cl I\to\map_{(2)}((X,A),(Y,B))$ from $f$ to $g$. However, we want $H$ to be a morphism in $\mathcal{T}_{(2)}$, but neither $I$ nor $\map_{(2)}((X,A),(Y,B))$ are objects in $\mathcal{T}_{(2)}$. We can remedy this by defining $H\cl f\simeq g$ to be a map
\bse
H\cl (I,\vn)\longrightarrow (\map_{(2)}((X,A),(Y,B)),\map_{\circ}(X,B))
\ese
such that $H(0)=f$ and $H(1)=g$.
\es

\bp
Let $X$ be a topological space, and let $x, y \in X$. Say that $x \sim y$ if $x$ and $y$ are in the same path component of $X$ (i.e., if there is a path $\alpha \cl I \to X$ with $\alpha(0) = x$ and $\alpha(1) = y$).
\ben[label=(\alph*)]
\item Show that $\sim$ is an equivalence relation.
\item Conclude that homotopy is an equivalence relation in $\mathcal{T}_{\circ}$ and in $\mathcal{T}_*$.
\een
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es

\bx
The set $[X, Y ]$ is a pointed set---what is the basepoint?
\ex

\bs
\es

\bp
Let $f \cl X \to Y$ and $g \cl Y \to Z$ be two maps in $\mathcal{T}_*$.
\ben[label=(\alph*)]
\item Show that if $x \sim x' \in X$, then $f(x) \sim f(x')$ in $Y$.
\item Define $\pi_0(f)$ to make $\pi_0$ a covariant functor from $\mathcal{T}_*$ to the category $\mathbf{Sets}_*$ of pointed sets.
\item Show that if $f \simeq g$ in $\mathcal{T}$, then $\pi_0(f) = \pi_0(g)$.
\een
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\item 
\een
\es

\bx
Show that if $F$ respects homotopy and $G$ is a homotopy functor, then $G \circ F$ is a homotopy functor.
\ex

\bs
\es

\bp
\ben[label=(\alph*)]
\item Show that the functors $- \times Z$, $ - \rtimes -$, $-\wedge Z$ and $\map_*(- ,- )$ respect homotopy in $\mathcal{T}_*$. 
\item Show that the functors $-\times Z$, $Z\rtimes -$ and $\map_{\circ}(-,-)$ respect homotopy in $\mathcal{T}_{\circ}$. 
\een
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es

\bp
\ben[label=(\alph*)]
\item Interpret the homotopy set $[ - , - ]$ in terms of $\pi_0$.
\item  Is $[ - , - ]$ functorial? Is it a homotopy functor? Keep in mind that there are two variables.
\een 
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es

\bx
Interpret $\pi_0$ in terms of $[ - , - ]$.
\ex

\bs
\es

\bp
Show that the definitions of free and pointed homotopy in terms of cylinders are equivalent to those given in Section 4.1.1.
\ep

\bs
\es

\bp
Using the definition of this section, show that
\ben[label=(\alph*)]
\item homotopy is an equivalence relation and 
\item if $f \simeq \bar{f}$ and $g \simeq \bar{g}$, then $g \circ f\simeq \bar{g}\circ \bar{f}$ in both the pointed and unpointed contexts.
\een
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es

\bx
Show that if $X$ is path-connected, then any two paths are freely homotopic. If the paths are pointed, then they are pointed homotopic.
\ex

\bs
\es

\bp
\ben[label=(\alph*)]
\item Suppose there are path homotopies $\alpha_1 \simeq \alpha_2$ and $\beta_1 \simeq \beta_2$. Show that $\alpha_1 * \beta_1 \simeq \alpha_2 * \beta_2$, assuming the concatenation is defined.
\item Show that $\alpha * \overleftarrow{\alpha}\simeq *$ is path homotopic to the constant path at $\alpha(0)$.

{\scshape Hint}. For (b), consider the path which follows $\alpha$ from time $0$ to time $2t$, then sits still until time $2-2t$, after which it follows $\overleftarrow{\alpha}$ until time $1$.
\een
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es

\bp
Let $p \cl I \to I$ be any path from $0$ to $1$.
\ben[label=(\alph*)]
\item Show that $p$ is path homotopic to $\id_I$.
\item Show that $\alpha \circ p$ is path homotopic to $\alpha$. 
\een
{\scshape Hint}. Look ahead to Section 4.2.1.
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es

\bp
\ben[label=(\alph*)]
\item Show that the paths $(\alpha * \beta) * \gamma$ and $\alpha * (\beta * \gamma)$ are path homotopic.
\item Let $\alpha$ be a path in $X$ from $x$ to $y$, and let $\boxed{x}$ and $\boxed{y}$ be the constant paths at $x$ and at $y$, respectively. Show that there are path homotopies $\alpha * \boxed{y} \simeq \alpha \simeq \boxed{x} * \alpha$.
\een
{\scshape Hint}. The paths in question are reparametrisations of each other.
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es

\bx
Show that the paths $(\alpha * \beta) * \gamma$ and $\alpha * (\beta * \gamma)$ are equal if and only if all three paths are constant.
\ex

\bs
\es

\bp
Define reparametrisation of homotopies, and show that there is always a homotopy-of-homotopies from $H$ to any reparametrisation of $H$.
\ep

\bs
\es

\bp
Let $H \cl f \simeq g$ be a homotopy.
\ben[label=(\alph*)]
\item Show that $H * \boxed{g} \simeq H \simeq \boxed{f} * H$.
\item Show that there is a homotopy-of-homotopies $H * \overleftarrow{H}\simeq\boxed{f}$.
\een
\ep

\bs
\ben[label=(\alph*)]
\item 
\item 
\een
\es
















